6.867 Problem Set 1

### 2015-10-02 pset1/paper.tex
done with 2

\end{document}

>>>>

\subsection{Minimizing SSE with gradient descent}
We attempted to find maximum-likelihood values of $\theta$ using unmodified gradient descent and \begin{tt}fmin\_bfgs\end{tt}. Unmodified gradient descent ended up performing very poorly on this problem; with $M=3$ and a convergence criteria of $10^{-8}$, 100,000 iterations (each iteration including one function call and one gradient call) were not enough for the algorithm to converge with a step size of 0.01. The best-performing step size we found was 0.06, which converged in 49,003 iterations. Larger step sizes (as small as 0.065) ended up diverging and giving us no useful results. Reducing the convergence criteria to $10^{-5}$ (at which point there was a visible discrepancy between the maximum-likelihood polynomials produced by the closed-form formula and that produced by gradient descent) did not impact performance significantly; with a step size of 0.06, the algorithm still required approximately 30,000 iterations to converge. Adjusting the starting guess did not significantly affect the performance of the algorithm.

\begin{tt}fmin\_bfgs\end{tt} performed much better on this problem. The algorithm required onnly 162 function calls and 27 gradient calls (again, with the $M=3$ case) to converge to a solution (which was visually indistinguishable from the optimal solution found by the closed-form formula). Plots depicting our results can be found in Figure 1.

\begin{figure}
\includegraphics[scale=0.4]{figure2_1.png}
\includegraphics[scale=0.4]{figure2_2.png}
\caption{The maximum-likelihood polynomials ($M=3$) produced by gradient descent. The sin function used to generate the data is in green, the polynomial produced by the closed-form formula in red, and the polynomial produced by gradient descent in blue. Note that in the left figure, the red curve is almost invisible, as it is mostly covered by the blue curve. The figure on the left corresponds to a convergence criterion of $10^{-8}$, while the one on the right corresponds to a convergence criterion of $10^{-5}$. The \texttt{fmin\_bfgs} function produced a figure similar to the one on the left.}
\end{figure}

Because the data set that we used was generated using an underlying sin function, we investigated the behavior of the algorithm when trigonometric functions were used to generate $\Phi$. We initially expected that the algorithm would generate a fit that was closer to the original sin function, but it actually did not. The result of this can be seen in Figure 2. Attempting to fit data in this way without knowing how it was generated also carries some disadvantages. Any function that we generate with trigonometric basis functions must necessarily be periodic with period 1, so the fit may not generalize well to values of $x$ outside the interval supplied. Furthermore, trigonometric functions will have a difficult time characterizing simple relationships, such as linear correlations.

\begin{figure}
\includegraphics[scale=0.4]{figure2_3.png}
\caption{The maximum-likelihood function produced using trigonometric basis functions ($M=3$). This function (blue) does not actually appear to be more faithful to the original sin function (green) than the maximum-likelihood degree-3 polynomial (red).}
\end{figure}
\end{document}
